---
title: "Assignment / Investigate Database Recovery"
author: "Liyang Song"
date: "3/28/2023"
output:
  pdf_document: default
  html_document:
    df_print: paged
  html_notebook: default
---

## Q1

Choose a database recovery problem or scenario (perhaps from work) and then propose a solution using the techniques described in Chapter 11 in the textbook. Briefly describe the technique, when it is appropriate to use and what recovery problem it solves.

**Problem**

During the inventory management, there should be one transaction dealing with adjusting inventories after merchandise sold. For example, after one completed purchase, the total sale of the merchandise should increase, while the corresponding inventory of the merchandise should be decreased. These steps should be accomplished within one transaction such that other customers and staffs can see the updated inventory.

Given log records of 2 transaction examples:

[$T_1$, start]

[$T_1$, sale, 599, 699]

[$T_1$, inventory, 40, 39]

[$T_1$, commit]

[$T_2$, start]

[$T_2$, sale, 699, 899]

[$T_2$, inventory, 39, 37]

[$T_2$, commit]

Transaction $T_1$ records one purchase, the sale amount increases by 100, and the inventory decreases by 1. And transaction $T_2$ records another purchase, the sale amount increases by 200, and the inventory decreases by 2. Suppose these two transactions are executed serially in the order $T_1$ followed by $T_2$.

Now let's see several log based recovery techniques that can be used to deal with system fails during these two transactions.

**Deferred Update Based Techniques (NO-UNDO/REDO)**

Deferred update will not execute update operations untill the transaction has entered partially committed state. So any transaction fails before the commit will not require a redo of the transaction, while any fails after the commit will require a redo, to make sure updates passed to data on the disk. AFIM in logs will be used to make the recovery.

**Immediate Update Based Techniques (UNDO/NO-REDO)**

Immediate update will execute update operations to data on the disk immediately, and before the transaction entered partially committed state. So any transaction fails before the commit will require a undo, while any fails after the commit will not require a redo, as the changes have been updated on the disk. BFIM in logs will be used to make the recovery.

**UNDO/REDO**

When changes are possibly not passed to the database after the commit, both undo and redo will be required. In such a case, both BFIM and AFIM in logs will be required to make the recovery.

**Solution to the problem**

>if the database is **deferred update** based:

if fail before $T_1$ commit: no redo or undo

if fail after $T_1$ commit, before $T_2$ commit: redo $T_1$

if fail after $T_2$ commit: redo $T_1$, $T_2$
        
>if the database is **immediate update** based:

if fail before $T_1$ commit: undo $T_1$

if fail after $T_1$ commit, before $T_2$ commit: undo $T_2$

if fail after $T_2$ commit: no redo or undo

>if the database is **immediate update based but commit before changes database**:

if fail before $T_1$ commit: undo $T_1$

if fail after $T_1$ commit, before $T_2$ commit: redo $T_1$, undo $T_2$

if fail after $T_2$ commit: redo $T_1$, $T_2$

## Q2

Using any of the SQLite database we have previously worked with, write an update that requires related modification of multiple tables and conduct those updates within a transaction. Test the code so that you show that the transaction works and write one test where the transaction fails and rolls back.

```{r remove_objects}
rm(list = ls())
```

```{r connect_to_sqlite}
require(RSQLite)
fpath <- getwd()
dbfile <- 'sakila.db'
dbcon <- dbConnect(RSQLite::SQLite(), paste0(fpath, '/', dbfile))
```

```{r transaction_definition}
# A transaction of payment processing to update all related tables
# To simplify, assume that the customer and staff has existed in the database

doTransfer <- function(dbcon, customer_id, staff_id, amount, payment_date, rental_date, inventory_id) {
    txnFailed = FALSE
    
    # check customer_id and staff_id exist in database
    
    customer_id_vector <- dbGetQuery(dbcon, 'SELECT customer_id FROM customer')
    if (!(customer_id %in% customer_id_vector$customer_id)) {
        return (FALSE)
    }
    
    staff_id_vector <- dbGetQuery(dbcon, 'SELECT staff_id FROM staff')
    if (!(staff_id %in% staff_id_vector$staff_id)) {
        return (FALSE)
    }
    
    dbExecute(dbcon, "BEGIN TRANSACTION")
    
    # update rental table to add rental info
    
    sql <- 'INSERT INTO rental(rental_date, inventory_id, customer_id, staff_id, last_update)
            VALUES (?, ?, ?, ?, ?)'
    ps <- dbSendStatement(dbcon, sql,
                        params = list(rental_date, inventory_id, customer_id, staff_id, format(Sys.time(), "%Y-%m-%d %H:%M:%S")))
    
    if (dbGetRowsAffected(ps) < 1) {
        txnFailed = TRUE
    }
    dbClearResult(ps)
    
    rental_id_df <- dbGetQuery(dbcon, 'SELECT MAX(rental_id) FROM rental')
    rental_id <- as.numeric(rental_id_df[[1]])
    
    # update payment table to add payment info
    
    sql <- 'INSERT INTO payment(customer_id, staff_id, rental_id, amount, payment_date, last_update)
            VALUES (?, ?, ?, ?, ?, ?)'
    ps <- dbSendStatement(dbcon, sql,
                        params = list(customer_id, staff_id, rental_id, amount, payment_date, format(Sys.time(), "%Y-%m-%d %H:%M:%S")))
    if (dbGetRowsAffected(ps) < 1) {
        txnFailed = TRUE
    }
    dbClearResult(ps)
    
    
    # update inventory table to remove inventory
    
    sql <- 'DELETE FROM inventory
            WHERE inventory_id = ?'
    ps <- dbSendStatement(dbcon, sql,
                        params = list(inventory_id))
    if (dbGetRowsAffected(ps) < 1) {
        txnFailed = TRUE
    }
    dbClearResult(ps)
    
    # commit transaction if no failure, otherwise rollback
    
    if (txnFailed == TRUE)
        dbExecute(dbcon, "ROLLBACK TRANSACTION")
    else
        dbExecute(dbcon, "COMMIT TRANSACTION")
    
        # return status; TRUE if successful; FALSE if failed
    
        return (!txnFailed)
}
```

```{r transaction_test_success_1}
# test the transaction with a set of valid data

customer_id <- 50
staff_id <- 2
amount <- 4.99
payment_date <- '2006-12-23 11:10:06'
rental_date <- payment_date
inventory_id <- 17

status <- doTransfer(dbcon, customer_id, staff_id, amount, payment_date, rental_date, inventory_id)

if (status == TRUE) {
    cat('Transfer successful')
} else {
    cat('Transfer failed')
}
```

```{r transaction_test_success_2}
# check updated table (the first record by DESC order)

dbGetQuery(dbcon, 'SELECT * FROM payment ORDER BY payment_id DESC LIMIT 10')
```

```{r transaction_test_success_3}
# check updated table (the first record by DESC order)

dbGetQuery(dbcon, 'SELECT * FROM rental ORDER BY rental_id DESC LIMIT 10')
```

```{r transaction_test_success_4}
dbGetQuery(dbcon, 'SELECT * FROM inventory WHERE inventory_id=17')
```

```{r transaction_test_fail}
# inventory_id=17 has been deleted from last transaction
# so this transaction would fail

customer_id <- 50
staff_id <- 2
amount <- 4.99
payment_date <- '2006-12-23 11:10:06'
rental_date <- payment_date
inventory_id <- 17

status <- doTransfer(dbcon, customer_id, staff_id, amount, payment_date, rental_date, inventory_id)

if (status == TRUE) {
    cat('Transfer successful')
} else {
    cat('Transfer failed')
}
```

```{r disconnect_to_DB}
dbDisconnect(dbcon)
```
