---
title: "Assignment / Query XML"
author: "Liyang Song"
date: "4/4/2023"
output:
  html_document:
    df_print: paged
  pdf_document: default
  html_notebook: default
---

## Q4

Load the XML with the DTD into R using validation.

```{r}
require(XML)
require(RSQLite)

xmlfn <- "CustomersAndOrders.xml"
xmlDOM <- xmlParse(file = xmlfn, validate = T)
```

## Q5

Execute an XPath expression that returns the names of all customers that do not live in "MA"?

```{r}
xpathEx <- "//Customer[FullAddress/Region != 'MA']/ContactName"
names <- xpathSApply(xmlDOM, xpathEx, xmlValue)
names
```

## Q6

Using the result returned in (5), count how many customers there are.

```{r}
length(names)
```

## Q7

Using a combination of R and XPath, calculate the total amount paid for freight for all orders placed by customer with ID "GREAL".

```{r}
xpathEx <- "sum(//Order[CustomerID = 'GREAL']/ShipInfo/Freight)"
freight <- xpathSApply(xmlDOM, xpathEx, xmlValue)
cat("Total freight for all orders placed by customer with ID 'GREAL': ", freight)
```

## Q8

Using a combination of R and XPath, calculate the average amount paid for freight for all orders shipped to the USA.

```{r}
xpathEx <- "//ShipInfo[ShipCountry = 'USA']/Freight"
freights <- xpathSApply(xmlDOM, xpathEx, xmlValue)
avg <- mean(as.numeric(freights))
cat("Average freight for all orders shipped to the USA: ", avg)
```

