---
title: "Song.CRDB.CS5200"
output: html_notebook
---

- Name: Liyang Song
- Course: CS5200 23Spring
- Date: 2/7/2023

The original ER Diagram has been edited to deal with many-to-many relationship and a lookup table was added. Check the new ERD [**here**](https://lucid.app/lucidchart/920be975-48a4-413f-b5ca-5c8410c99419/edit?viewport_loc=-304%2C-86%2C1917%2C774%2COx9uyAtKxg0w&invitationId=inv_833a8424-fed1-49a0-9c2a-05785a462e07).

## 1. Create database file and initiate database connection.

```{r}
library(RSQLite)

fpath = "C:/users/Leon/documents/NEU/CS5200/Assignment4/"
dbfile = "ModuleAuthor.sqlite"

dbcon <- dbConnect(RSQLite::SQLite(), paste0(fpath, dbfile))

dbExecute(dbcon, "PRAGMA foreign_keys = ON")
```

## 2. Create table **Module**.

```{sql connection=dbcon}
DROP TABLE IF EXISTS Module;
```

```{sql connection=dbcon}
CREATE TABLE Module(
  mID TEXT,
  title TEXT NOT NULL,
  lengthInMin INTEGER NOT NULL,
  PRIMARY KEY (mID)
);
```

## 3. Create table **Affiliation**

```{sql connection=dbcon}
DROP TABLE IF EXISTS Affiliation
```

```{sql connection=dbcon}
CREATE TABLE Affiliation(
  affID INTEGER,
  name TEXT NOT NULL,
  PRIMARY KEY (affID)
)
```

## 4. Create table **Author**

```{sql connection=dbcon}
DROP TABLE IF EXISTS Author;
```

```{sql connection=dbcon}
CREATE TABLE Author(
  aid INTEGER,
  name TEXT NOT NULL,
  email TEXT NOT NULL,
  affID INTEGER,
  PRIMARY KEY (aid),
  FOREIGN KEY (affID) REFERENCES Affiliation (affID) ON DELETE SET NULL ON UPDATE CASCADE
);
```

## 5. Create table **ModuleAuthor**

```{sql connection=dbcon}
DROP TABLE IF EXISTS ModuleAuthor;
```

```{sql connection=dbcon}
CREATE TABLE ModuleAuthor(
  maID INTEGER,
  mID TEXT NOT NULL,
  aID INTEGER NOT NULL,
  PRIMARY KEY (maID),
  FOREIGN KEY (mID) REFERENCES Module (mID) ON DELETE CASCADE ON UPDATE CASCADE,
  FOREIGN KEY (aID) REFERENCES Author (aid) ON DELETE CASCADE ON UPDATE CASCADE
);
```

## 6. Fill tables with test data

```{sql connection=dbcon}
INSERT INTO Module VALUES
  ("CS5200", "Database Management Systems", 200),
  ("DS5110", "Intro to Data Management", 180),
  ("CS5800", "Algorithms", 300);
```

```{sql connection=dbcon}
INSERT INTO Affiliation VALUES
  (1, "PT"),
  (2, "FT"),
  (3, "Contract");
```

```{sql connection=dbcon}
INSERT INTO Author VALUES
  (1, "David", "david@northeastern.edu", 1),
  (2, "Sophia", "sophia@northeastern.edu", 1),
  (3, "Allen", "allen@northeastern.edu", 2);
```

```{sql connection=dbcon}
INSERT INTO ModuleAuthor VALUES
  (1, "CS5200", 3),
  (2, "DS5110", 3),
  (3, "CS5800", 1);
```

## 7. Check table contents

```{sql connection=dbcon}
SELECT * FROM Module;
```

```{sql connection=dbcon}
SELECT * FROM Affiliation;
```

```{sql connection=dbcon}
SELECT * FROM Author;
```

```{sql connection=dbcon}
SELECT * FROM ModuleAuthor;
```

## 8. Test extensibility of affiliation

```{sql connection=dbcon}
INSERT INTO Affiliation VALUES
  (4, "Intern"),
  (5, "Third-Party");
```

```{sql connection=dbcon}
SELECT * FROM Affiliation;
```

## 9. Test behavior for foreign keys

**SET NULL behavior**

```{sql connection=dbcon}
DELETE FROM Affiliation WHERE affID = 2;
```

```{sql connection=dbcon}
SELECT * FROM Affiliation;
```

```{sql connection=dbcon}
SELECT * FROM Author;
```

**CASCADE behavior**

```{sql connection=dbcon}
DELETE FROM Module WHERE mID = "CS5800";
```

```{sql connection=dbcon}
SELECT * FROM Module;
```

```{sql connection=dbcon}
SELECT * FROM ModuleAuthor;
```

## 10. Close database connection

```{r}
dbDisconnect(dbcon)
```
