---
title: "Song.DBTriggers.CS5200"
output: html_notebook
---

- Name: Liyang Song
- Course: CS5200 23Spring
- Date: 2/20/2023

The original ER Diagram has been edited to deal with many-to-many relationship, and a lookup table and a log table were added. Check the new ERD [**here**](https://lucid.app/lucidchart/3b1cb7f4-0d65-4165-b960-0625fb41c9ce/edit?viewport_loc=-296%2C-125%2C1277%2C516%2COx9uyAtKxg0w&invitationId=inv_37daab7e-c1ad-4d95-89d2-a5bd8c363d64).

## 1. Create database file and initiate database connection.

```{r}
library(RSQLite)

fpath = getwd()
dbfile = "DBTriggers.sqlite"

dbcon <- dbConnect(RSQLite::SQLite(), paste0(fpath, "/", dbfile))

dbExecute(dbcon, "PRAGMA foreign_keys = ON")
```

## 2. Create table **Module**, **Affiliation**, **Author**, **ModuleAuthor**, **Log**.

### Module

```{sql connection=dbcon}
DROP TABLE IF EXISTS Module;
```

```{sql connection=dbcon}
CREATE TABLE IF NOT EXISTS Module(
  mID TEXT,
  title TEXT NOT NULL,
  lengthInMin INTEGER NOT NULL,
  PRIMARY KEY (mID)
);
```

### Affiliation

```{sql connection=dbcon}
DROP TABLE IF EXISTS Affiliation;
```

```{sql connection=dbcon}
CREATE TABLE IF NOT EXISTS Affiliation(
  affID INTEGER PRIMARY KEY AUTOINCREMENT,
  name TEXT NOT NULL
);
```

### Author

```{sql connection=dbcon}
DROP TABLE IF EXISTS Author;
```

```{sql connection=dbcon}
CREATE TABLE IF NOT EXISTS Author(
  aid INTEGER PRIMARY KEY AUTOINCREMENT,
  name TEXT NOT NULL,
  email TEXT NOT NULL,
  affID INTEGER,
  FOREIGN KEY (affID) REFERENCES Affiliation (affID) ON DELETE SET NULL ON UPDATE CASCADE
);
```

### ModuleAuthor

```{sql connection=dbcon}
DROP TABLE IF EXISTS ModuleAuthor;
```

```{sql connection=dbcon}
CREATE TABLE IF NOT EXISTS ModuleAuthor(
  maID INTEGER PRIMARY KEY AUTOINCREMENT,
  mID TEXT NOT NULL,
  aID INTEGER NOT NULL,
  FOREIGN KEY (mID) REFERENCES Module (mID) ON DELETE CASCADE ON UPDATE CASCADE,
  FOREIGN KEY (aID) REFERENCES Author (aid) ON DELETE CASCADE ON UPDATE CASCADE
);
```

### Log

```{sql connection=dbcon}
DROP TABLE IF EXISTS Log
```

```{sql connection=dbcon}
CREATE TABLE IF NOT EXISTS Log(
    ID INTEGER PRIMARY KEY AUTOINCREMENT,
    action TEXT NOT NULL,
    DT TEXT NOT NULL
);
```

## 3. Create trigger **InsertLogs**, **UpdateLogs**, **DeleteLogs**, **AuthorNumPerModule**.

### InsertLogs

Trigger `InsertLogs` will add a record of any insertion to table `ModuleAuthor`. The recorded information includes `maID`, `mID`, `aID`, as well as a time stamp.

```{sql connection=dbcon}
DROP TRIGGER IF EXISTS InsertLogs
```

```{sql connection=dbcon}
CREATE TRIGGER IF NOT EXISTS InsertLogs
AFTER INSERT ON ModuleAuthor
  FOR EACH ROW
  BEGIN
    INSERT INTO Log(action, DT)
    VALUES (("Insert maID="||NEW.maID||" with mID="||NEW.mID||" and aID="||NEW.aID), DATETIME('NOW'));
  END;
```

### UpdateLogs

Trigger `UpdateLogs` will add a record of any update to table `ModuleAuthor`. The recorded information includes `maID`, `mID`, `aID`, as well as a time stamp.

```{sql connection=dbcon}
DROP TRIGGER IF EXISTS UpdateLogs
```

```{sql connection=dbcon}
CREATE TRIGGER IF NOT EXISTS UpdateLogs
AFTER UPDATE ON ModuleAuthor
  FOR EACH ROW
  BEGIN
    INSERT INTO Log(action, DT)
    VALUES (("Update maID="||NEW.maID||" from mID="||OLD.mID||" and aID="||OLD.aID||" to mID="||NEW.mID||" and aID="||NEW.aID), DATETIME('NOW'));
  END;
```

### DeleteLogs

Trigger `DeleteLogs` will add a record of any deletion to table `ModuleAuthor`. The recorded information includes `maID`, `mID`, `aID`, as well as a time stamp.

```{sql connection=dbcon}
DROP TRIGGER IF EXISTS DeleteLogs
```

```{sql connection=dbcon}
CREATE TRIGGER IF NOT EXISTS DeleteLogs
AFTER DELETE ON ModuleAuthor
  FOR EACH ROW
  BEGIN
    INSERT INTO Log(action, DT)
    VALUES (("Delete maID="||OLD.maID||" with mID="||OLD.mID||" and aID="||OLD.aID), DATETIME('NOW'));
  END;
```

### AuthorNumPerModule

Trigger `AuthorNumPerModule` will set a limit on the total number of authors per module in table `ModuleAuthor`.

When a new insertion occurs and the total number of authors per module will exceed the limit, any extra value will be deleted. As a result, only limited number of values will be shown in the table `ModuleAuthor`, yet all actions including insertion and deletion can be recorded in the table `Log`.

```{sql connection=dbcon}
DROP TRIGGER IF EXISTS AuthorNumPerModule
```

```{sql connection=dbcon}
CREATE TRIGGER IF NOT EXISTS AuthorNumPerModule AFTER INSERT ON ModuleAuthor
  FOR EACH ROW
  WHEN (SELECT COUNT(DISTINCT aID) FROM ModuleAuthor WHERE mID = NEW.mID) > 3
  BEGIN
    DELETE FROM ModuleAuthor
    WHERE maID = NEW.maID;
  END;
```

## 4. Check database triggers.

```{sql connection=dbcon}
SELECT * FROM sqlite_master WHERE TYPE = 'trigger';
```

## 5. Fill tables with test data and check table contents.

### Module

```{sql connection=dbcon}
INSERT INTO Module VALUES
  ("CS5200", "Database Management Systems", 200),
  ("DS5110", "Intro to Data Management", 180),
  ("CS5800", "Algorithms", 300);
```

```{sql connection=dbcon}
SELECT * FROM Module;
```

### Affiliation

```{sql connection=dbcon}
INSERT INTO Affiliation (name) VALUES
  ("PT"),
  ("FT"),
  ("Contract");
```

```{sql connection=dbcon}
SELECT * FROM Affiliation;
```

### Author

```{sql connection=dbcon}
INSERT INTO Author (name, email, affID) VALUES
  ("David", "david@northeastern.edu", 1),
  ("Sophia", "sophia@northeastern.edu", 1),
  ("Allen", "allen@northeastern.edu", 2),
  ("Eric", "eric@northeastern.edu", 3),
  ("Natalia", "natalia@northeastern.edu", 3);
```

```{sql connection=dbcon}
SELECT * FROM Author;
```

### ModuleAuthor

```{sql connection=dbcon}
INSERT INTO ModuleAuthor (mID, aID) VALUES
  ("CS5200", 3),
  ("DS5110", 3),
  ("CS5800", 1);
```

```{sql connection=dbcon}
SELECT * FROM ModuleAuthor;
```

### Log

```{sql connection=dbcon}
SELECT * FROM Log;
```

## 6. Test trigger **InsertLogs**.

A test with a new insertion to the `ModuleAuthor`. The information of this action was recorded in the `Log` successfully.

```{sql connection=dbcon}
INSERT INTO ModuleAuthor (mID, aID) VALUES
  ("DS5110", 4);
```

```{sql connection=dbcon}
SELECT * FROM ModuleAuthor;
```

```{sql connection=dbcon}
SELECT * FROM Log;
```

## 7. Test trigger **UpdateLogs**.

A test with a new update to the `ModuleAuthor`. The information of this action was recorded in the `Log` successfully.

```{sql connection=dbcon}
UPDATE ModuleAuthor
SET aID=2
WHERE mID="CS5800"
```

```{sql connection=dbcon}
SELECT * FROM ModuleAuthor;
```

```{sql connection=dbcon}
SELECT * FROM Log;
```

## 8. Test trigger **DeleteLogs**.

A test with a new deletion to the `ModuleAuthor`. The information of this action was recorded in the `Log` successfully.

```{sql connection=dbcon}
DELETE FROM ModuleAuthor
WHERE mID="CS5200"
```

```{sql connection=dbcon}
SELECT * FROM ModuleAuthor;
```

```{sql connection=dbcon}
SELECT * FROM Log;
```

## 8. Test trigger **AuthorNumPerModule**.

A test with several new insertion to the `ModuleAuthor`.

**Four** authors were added to associate with module "CS5200", yet only the first three ones were inserted. The reason is that we set a limit of author number per module as **three**. And the forth one was inserted but then deleted by the trigger.

Meanwhile, the information of all actions was recorded in the `Log` successfully.

```{sql connection=dbcon}
INSERT INTO ModuleAuthor (mID, aID) VALUES
  ("CS5200", 1),
  ("CS5200", 2),
  ("CS5200", 4),
  ("CS5200", 5);
```


```{sql connection=dbcon}
SELECT * FROM ModuleAuthor;
```

```{sql connection=dbcon}
SELECT * FROM Log;
```

## 9. Close database connection.

```{r}
dbDisconnect(dbcon)
```
